A forest is more than its trees: haplotypes and inferred ARGs

### 2024-10-12 paper/writeup.tex
more on tree metrics

Haplotype identity has been relatively overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identiical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure sharing of haplotypes between adjacent trees.

@@

If we begin with a tree sequence containing unary nodes,
it is straightforward to remove the portions of each node's span on which it is unary,
apply \algorithmref{alg:extend},
and quantify how much node span was correctly or incorrectly added.
However, we are also interested in whether \algorithmref{alg:extend}
improves \emph{inferred} tree sequences.
Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
that take into account haplotype identity,
we define a measure of \emph{discrepancy} to quantify this.

@@

\comment{PREVIOUS VERSION: retrieve bits from below?}

@@

>>>>

Haplotype identity has been nearly overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identiical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure sharing of haplotypes between adjacent trees.

@@

If we begin with a tree sequence containing unary nodes,
it is straightforward to remove the portions of each node's span on which it is unary,
apply \algorithmref{alg:extend},
and quantify how much node span was correctly or incorrectly added.
However, we are also interested in whether \algorithmref{alg:extend}
improves \emph{inferred} tree sequences.
Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
that take into account haplotype identity,
we define a measure of \emph{discrepancy} to quantify this.
The method is implemented in the \texttt{tsdate} package \citep{wohns2022unified}
as \comment{XXX}.

@@

%%%%%%%%%%%%
\comment{PREVIOUS VERSION: retrieve bits from below?}

@@

\paragraph{Metrics on ARGs}
\comment{collecting random notes here; TODO is tie them together}
The discrepancy $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
This is by design: in practice it is not possible to infer all aspects
of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
and so we wanted to quantify
``How much of the true relationships does this ARG represent?''
However, the symmetrized version
$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
and view each ARG as a subset of the space $[0,L) \times N \times N$,
where $N$ is the shared set of nodes.
Then, discrepancy is the Lesbegue measure of the relative difference of the two sets:
$|\T_1 \setminus \T_2|$,
and so the symmetrized version is the measure of the symmetric difference,
which is well-known to be a metric.
If the two ARGs have the same number of nodes,
then we take the minimum over the all bijections between their nodes,
and minima over a finite number of metrics is also a metric.
Finally, if they have a different number of nodes,
then \comment{???}.

The Robinson-Foulds distance \citep{robinson1981comparison}
essentially counts the number of differing branches between two trees;
the averaged Robinson-Foulds distance \citep{kelleher2019inferring} % TODO: first usage?
combines this distance across marginal trees, weighted by span along the genome.
The method we present here for measuring discrepancy between topologies of ARGs
is a straightforward generalization
that takes into account span along the genome of inferred ancestral haplotypes
(and separates the metric into two pieces as discussed in the previous paragraph).
However, the Robinson-Foulds metric has many undesireable properties --
for instance, moving a single tip can result in a tree with maximum distance to the original --
and there is a substantial literature giving generalizations
\citep[reviewed by][]{llabres2021generalized}.
Many of these generalizations \citep[e.g.,][]{bocker2013generalized}
relax the requirement that the match between subtended sample sets be exact,
and weight matches in some way by the size of the discrepancy.
We considered such definitions as well, but kept to the simple case
for computational tractability --
the generalization of \citet{bocker2013generalized} is NP-hard to compute, even for a single tree.
In the ARG literature, \citet{zhang2023biobankscale}
defines a metric (called ``ARG total variation distance'') that includes branch lengths,
in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
however, it is still applied to ARGs as an average over marginal trees,
without enforcement of identity across haplotypes;
it would be useful to extend our discrepancy to include branch lengths.
